Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-09-14 source-iclr/app_sumarization.tex
add scaffold and analysis

\section{Comparison of convergence rates with related works.}
\label{sec:app:comparison}
In this section, we compare our convergence rate with the best-known results in the literature (see Table~\ref{tb:convergenceratev3}). 
% In Table~\ref{tb:convergenceratev1} and Table~\ref{tb:convergenceratev2}, we summarize 
% the convergence results in terms of participations and accelerations. 
% In Table~\ref{tb:convergenceratev3}, we compare our 
% results with the best known results in the literatures. 
In~\cite{haddadpour2019convergence}, the authors provide $\cO(1/NT)$ convergence rate of non-convex problems under Polyak-Łojasiewicz (PL) condition, which
means their results can directly apply to the strongly convex problems. However, their assumption is based on bounded gradient diversity, defined as follows: 
\begin{align*}
	\Lambda(\vw) = \frac{\sum_{k}p_{k}\|\nabla F_{k}(\mathbf{w})\|_{2}^{2}}{\|\sum_{k}p_{k}\nabla F_{k}(\mathbf{w})\|_{2}^{2}} \leq B
\end{align*} 
This is a more restrictive assumption comparing to assuming bounded gradient under the case of target accuracy $\epsilon \rightarrow 0$ and PL condition.
To see this, consider the gradient diversity at the global optimal $\vw^*$, i.e., $\Lambda(\vw^*) = \frac{\sum_{k}p_{k}\|\nabla F_{k}(\mathbf{w})\|_{2}^{2}}{\|\sum_{k}p_{k}\nabla F_{k}(\mathbf{w})\|_{2}^{2}}$. For $\Lambda(\vw^*)$ to be bounded, it requires $\|\nabla F_{k}(\mathbf{w}^*)\|_{2}^{2} = 0$, $\forall \ k$. This indicates 
$\vw^*$ is also the minimizer of each local objective, which contradicts to the practical setting of heterogeneous data. Therefore, their bound 
is not effective for arbitrary small $\epsilon$-accuracy under general heterogeneous data while our convergence results still hold in this case.
% If we only consider the domain $\vw \in \{\vw | \|\sum_{k}p_{k}\nabla F_{k}(\mathbf{w})\|_{2}^{2} > c\}$, where $c$ is a positive constant. 
% The bounded gradient can imply gradient diversity (see Corollary 10 in ~\cite{li2018federated}). In this case, the gradient diversity is a relaxed

@@

\begin{table}[h!]
\centering
{\tiny
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline Reference                 & Convergence rate    & E                           			& NonIID & Participation & Extra Assumptions  		  & Setting  \\ \hline\hline 
FedAvg\cite{li2019convergence}         & $\cO(\frac{E^2}{T})$& $\cO(1)$                    		& \cmark & Partial       & Bounded gradient   		  & Strongly convex  \\ \hline
FedAvg\cite{haddadpour2019convergence} & $\cO(\frac{1}{KT})$ & $\cO(K^{-1/3}T^{2/3})^{\dagger}$ & \cmark$^{\ddagger\ddagger}$ & Partial       & Bounded gradient diversity   & Strongly convex$^{\mathsection}$  \\ \hline
FedAvg\cite{koloskova2020unified} & $\cO(\frac{1}{NT})$ & $\cO(N^{-1/2}T^{1/2})$     	& \cmark & Full       & Bounded gradient   & Strongly convex  \\ \hline
FedAvg/N-FedAvg                 & $\cO(\frac{1}{KT})$ & $\cO(N^{-1/2}T^{1/2})^{\ddagger}$ & \cmark	 & Partial       & Bounded gradient             & Strongly convex  \\\hline\hline
FedAvg\cite{khaled2020tighter}  & $\cO(\frac{1}{\sqrt{NT}})$ & $\cO(N^{-3/2}T^{1/2})$     	    & \cmark& Full        & Bounded gradient             & Convex  \\\hline
FedAvg\cite{koloskova2020unified} & $\cO(\frac{1}{\sqrt{NT}})$ & $\cO(N^{-3/4}T^{1/4})$    & \cmark & Full       & Bounded gradient             &  Convex  \\ \hline
FedAvg/N-FedAvg      & $\cO\left(\frac{1}{\sqrt{KT}}\right)$ & $\cO(N^{-3/4}T^{1/4})^{\ddagger}$& \cmark			& Partial     & Bounded gradient            &  Convex   \\ \hline\hline
FedAvg & $\cO\left(\exp(-\frac{NT}{E\kappa_1})\right)$ & $ \cO(T^{\beta})$                   & \cmark&  Partial     & Bounded gradient    & Overparameterized LR\\ \hline
FedMass & $\cO\left(\exp(-\frac{NT}{E\sqrt{\kappa_1\tilde{\kappa}}})\right)$ & $ \cO(T^{\beta})$ & \cmark &  Partial     & Bounded gradient    & Overparameterized LR \\ \hline
\end{tabular}
}
\caption{A high-level summary of the convergence results in this paper compared to prior state-of-the-art FL algorithms. This table only highlights the
dependence on $T$ (number of iterations), $E$ (the maximal number of local steps), $N$ (the total number of devices), and $K\leq N$ the number of participated devices. 
$\kappa$ is the condition number of the system and $\beta \in (0,1)$. We denote Nesterov accelerated FedAvg as N-FedAvg in this table.}
{\raggedright 
         $^{\dagger}$ This $E$ is obtained under i.i.d. setting. \\
         $^{\ddagger}$ This $E$ is obtained under full participation setting. \\ 
         % $^{\ddagger}$ The convergence of FedAvg for general overparameterized strongly convex and overparameterized linear regression share the same order in terms of $E, T, N$.\\
         $^{\mathsection}$ In~\cite{haddadpour2019convergence}, the convergence rate is for non-convex smooth problems with PL condition, which also applies to strongly convex problems. Therefore, we compare it with our strongly convex results here.\\
         $^{\ddagger\ddagger}$ The bounded gradient diversity assumption is not applicable for general heterogeneous data when converging to arbitrarily small $\epsilon$-accuracy (see discussions in Sec~\ref{sec:app:comparison}).
           \par}
\label{tb:convergenceratev3}
\end{table}

>>>>

\section{Comparison of Convergence Rates with Related Works}
\label{sec:app:comparison}
In this section, we compare our convergence rate with the best-known results in the literature (see Table~\ref{tb:convergenceratev3}). 
% In Table~\ref{tb:convergenceratev1} and Table~\ref{tb:convergenceratev2}, we summarize 
% the convergence results in terms of participations and accelerations. 
% In Table~\ref{tb:convergenceratev3}, we compare our 
% results with the best known results in the literatures. 
In~\cite{haddadpour2019convergence}, the authors provide $\cO(1/NT)$ convergence rate of non-convex problems under Polyak-Łojasiewicz (PL) condition, which
means their results can directly apply to the strongly convex problems. However, their assumption is based on bounded gradient diversity, defined as follows: 
\begin{align*}
	\Lambda(\vw) = \frac{\sum_{k}p_{k}\|\nabla F_{k}(\mathbf{w})\|_{2}^{2}}{\|\sum_{k}p_{k}\nabla F_{k}(\mathbf{w})\|_{2}^{2}} \leq B
\end{align*} 
This is a more restrictive assumption comparing to assuming bounded gradient under the case of target accuracy $\epsilon \rightarrow 0$ and PL condition.
To see this, consider the gradient diversity at the global optimal $\vw^*$, i.e., $\Lambda(\vw^*) = \frac{\sum_{k}p_{k}\|\nabla F_{k}(\mathbf{w})\|_{2}^{2}}{\|\sum_{k}p_{k}\nabla F_{k}(\mathbf{w})\|_{2}^{2}}$. For $\Lambda(\vw^*)$ to be bounded, it requires $\|\nabla F_{k}(\mathbf{w}^*)\|_{2}^{2} = 0$, $\forall \ k$. This indicates 
$\vw^*$ is also the minimizer of each local objective, which contradicts to the practical setting of heterogeneous data. Therefore, their bound 
is not effective for arbitrary small $\epsilon$-accuracy under general heterogeneous data while our convergence results still hold in this case.
% If we only consider the domain $\vw \in \{\vw | \|\sum_{k}p_{k}\nabla F_{k}(\mathbf{w})\|_{2}^{2} > c\}$, where $c$ is a positive constant. 
% The bounded gradient can imply gradient diversity (see Corollary 10 in ~\cite{li2018federated}). In this case, the gradient diversity is a relaxed

In~\cite{karimireddy2019scaffold}, the linear speedup convergence rate of FedAvg are provided for strongly convex, general convex, and non-convex problems 
under full participation setting. However, their rate does not enjoy linear speedup for any number of devices while our results apply to any valid $K\leq N$. For example,  they provides an optimality gap of $\cO\left((1-\frac{K}{N})E/T\right)$ for the strongly convex case \cite[Theorem V]{karimireddy2019scaffold}.
With partial participation, and when $K=O(1)$, their convergence rate is $\cO(E/T)$ which does not have linear speedup. 
Under partial participation, the FedAvg analyses in~\cite{karimireddy2019scaffold} requires $E=\cO(1)$. For example, the strongly convex result $\cO((1-\frac{K}{N})E/T)$ in Theorem V is $\cO(E/T)$ when $K=O(1)$ and is $\cO(E/NT)$ when $K=\cO(N)$. In either case, to achieve a $\cO(1/T)$ convergence rate, it requires $E=\cO(1)$ as well. Similar conclusion also holds for the general convex problem.

@@

\begin{table}[h!]
\hspace{-4em}
{
\centering
\tiny
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline Reference                 & Convergence rate    & E                           			& NonIID & Participation & Extra Assumptions  		  & Setting  \\ \hline\hline 
FedAvg\cite{li2019convergence}         & $\cO(\frac{E^2}{T})$& $\cO(1)$                    		& \cmark & Partial       & Bounded gradient   		  & Strongly convex  \\ \hline
FedAvg\cite{haddadpour2019convergence} & $\cO(\frac{1}{KT})$ & $\cO(K^{-1/3}T^{2/3})^{\dagger}$ & \cmark$^{\ddagger\ddagger}$ & Partial       & Bounded gradient diversity   & Strongly convex$^{\mathsection}$  \\ \hline
FedAvg\cite{koloskova2020unified} & $\cO(\frac{1}{NT})$ & $\cO(N^{-1/2}T^{1/2})$     	& \cmark & Full       & Bounded gradient   & Strongly convex  \\ \hline
FedAvg\cite{karimireddy2019scaffold} & $\cO(\frac{1}{NT})^{\dagger\dagger}$ & $\cO(1)^{\dagger\dagger}$   	& \cmark & Partial       & Bounded gradient dissimilarity   & Strongly convex  \\ \hline
FedAvg/N-FedAvg                 & $\cO(\frac{1}{KT})$ & $\cO(N^{-1/2}T^{1/2})^{\ddagger}$ & \cmark	 & Partial       & Bounded gradient             & Strongly convex  \\\hline\hline
FedAvg\cite{khaled2020tighter}  & $\cO(\frac{1}{\sqrt{NT}})$ & $\cO(N^{-3/2}T^{1/2})$     	    & \cmark& Full        & Bounded gradient             & Convex  \\\hline
FedAvg\cite{koloskova2020unified} & $\cO(\frac{1}{\sqrt{NT}})$ & $\cO(N^{-3/4}T^{1/4})$    & \cmark & Full       & Bounded gradient             &  Convex  \\ \hline
FedAvg\cite{karimireddy2019scaffold} & $\cO(\frac{1}{\sqrt{NT}})^{\dagger\dagger}$ & $\cO(1)^{\dagger\dagger}$  & \cmark & Partial       & Bounded gradient dissimilarity   &  Convex  \\ \hline
FedAvg/N-FedAvg      & $\cO\left(\frac{1}{\sqrt{KT}}\right)$ & $\cO(N^{-3/4}T^{1/4})^{\ddagger}$& \cmark			& Partial     & Bounded gradient            &  Convex   \\ \hline\hline
FedAvg & $\cO\left(\exp(-\frac{NT}{E\kappa_1})\right)$ & $ \cO(T^{\beta})$                   & \cmark&  Partial     & Bounded gradient    & Overparameterized LR\\ \hline
FedMass & $\cO\left(\exp(-\frac{NT}{E\sqrt{\kappa_1\tilde{\kappa}}})\right)$ & $ \cO(T^{\beta})$ & \cmark &  Partial     & Bounded gradient    & Overparameterized LR \\ \hline
\end{tabular}
}
\caption{A high-level summary of the convergence results in this paper compared to prior state-of-the-art FL algorithms. This table only highlights the
dependence on $T$ (number of iterations), $E$ (the maximal number of local steps), $N$ (the total number of devices), and $K\leq N$ the number of participated devices. 
$\kappa$ is the condition number of the system and $\beta \in (0,1)$. We denote Nesterov accelerated FedAvg as N-FedAvg in this table.}
{\raggedright 
         $^{\dagger}$ This $E$ is obtained under i.i.d. setting. \\
         $^{\ddagger}$ This $E$ is obtained under full participation setting. \\ 
         % $^{\ddagger}$ The convergence of FedAvg for general overparameterized strongly convex and overparameterized linear regression share the same order in terms of $E, T, N$.\\
         $^{\mathsection}$ In~\cite{haddadpour2019convergence}, the convergence rate is for non-convex smooth problems with PL condition, which also applies to strongly convex problems. Therefore, we compare it with our strongly convex results here.\\
         $^{\ddagger\ddagger}$ The bounded gradient diversity assumption is not applicable for general heterogeneous data when converging to arbitrarily small $\epsilon$-accuracy (see discussions in Sec~\ref{sec:app:comparison}).\\
         $^{\dagger\dagger}$ Although the results in~\cite{karimireddy2019scaffold} is applicable for partial participation setting, their results only achieve linear speedup under full participation setting $K=N$ while we show linear speedup convergence for $K\leq N$ (see discussions in Sec~\ref{sec:app:comparison}). Under partial participation, the communication complexity is $\cO(1)$.)
           \par}
\label{tb:convergenceratev3}
\end{table}

### 2020-09-14 source-iclr/sec_intro.tex
add scaffold and analysis

Despite its promise and broad applicability in our current era, the potential value FL delivers is coupled with the unique challenges it brings forth. In particular, when FL learns a single statistical model using data from across all the devices while keeping each individual device's data isolated (and hence protects privacy)~\cite{kairouz2019advances}, it faces two challenges that are absent in centralized optimization and distributed (stochastic) optimization~\cite{zhou2017convergence, stich2018local,khaled2019first,liang2019variance,wang2018cooperative,woodworth2018graph,wang2019adaptive,jiang2018linear,yu2019parallel,yu2019linear, khaled2020tighter,koloskova2020unified}:

>>>>

Despite its promise and broad applicability in our current era, the potential value FL delivers is coupled with the unique challenges it brings forth. In particular, when FL learns a single statistical model using data from across all the devices while keeping each individual device's data isolated~\cite{kairouz2019advances}, it faces two challenges that are absent in centralized optimization and distributed (stochastic) optimization~\cite{zhou2017convergence, stich2018local,khaled2019first,liang2019variance,wang2018cooperative,woodworth2018graph,wang2019adaptive,jiang2018linear,yu2019parallel,yu2019linear, khaled2020tighter,koloskova2020unified}:
